### 2020-02-03 ncov.tex
ok
update

We evaluate six disparate models published online between January 24--26, 2020 that estimated \Ro for the 2019-nCoV outbreak \citep{imaincov, liuncov, majumderncov, readncov, riouncov, zhaoncov}. 
We use our framework to construct pooled estimates for the three key quantities: $r$, $\bar G$, and $\kappa$;
the pooled estimates reflect the uncertainties present in modeling approaches and their underlying assumptions.
We use these pooled estimates to illustrate the importance of propagating different sources of error, particularly uncertainty in both the growth rate and the generation interval. 
We also use our framework to tease apart which assumptions of these different models led to their different estimates and confidence intervals.

@@

\begin{table}[t]
\begin{center}
\scriptsize
\begin{tabular}{l|p{3.5cm}|p{2.5cm}|p{2.7cm}|l}
 & Basic reproductive\newline number \Ro & Mean generation\newline interval $\bar G$ (days) & Generation-interval\newline dispersion $\kappa$ \\
\hline
Study 1 & 2.5 (1.5--3.5)$^\ast$ & 8.4 & unspecified$^\dagger$ & \cite{imaincov} \\
\hline
Study 2 & 2.92 (95\% CI: 2.28--3.67) & 8.4 & 0.2 & \cite{liuncov} \\
\hline
Study 3 & 3.8 (95\% CI: 3.6--4.0) & 7.6 & 0.5 & \cite{readncov} \\
\hline
Study 4 & 2.2 (90\% CI: 1.4--3.8) & 7--14 & 0.5 & \cite{riouncov} \\
\hline
Study 5 & 5.47 (95\% CI: 4.16--7.10)$^\ddagger$ & 7.6--8.4 & 0.2 & \cite{zhaoncov} \\
\hline
Study 6 & 2.0--3.1 & 6--10 & 0 & \cite{majumderncov} \\
\hline
\end{tabular}
\end{center}
\caption{
\textbf{Reported estimates of the basic reproductive number and the assumptions about the generation-interval distributions.}
Estimates of \Ro and their assumptions about the shape of the generation interval distributions were collected from 6 studies.
$^\ast$We treat these intervals as a 95\% confidence interval in our analysis.
$^\dagger$We assume $\kappa = 0.5$ in our analysis.
$^\ddagger$The authors presented \Ro estimates under different assumptions; we use their baseline scenario in our analysis.
}
\end{table}

We gathered information on estimates of \Ro and their assumptions about the underlying generation-interval distributions from 6 articles that were published online between January 24 and January 26, 2020 (Table 1).
As most studies do not report their estimates of the exponential growth rate or the associated confidence intervals, we first recalculate the exponential growth rate that correspond their model assumptions.
We do so by modeling explicitly or implicitly reported distributions of the reproductive number \Ro, the mean generation interval $\bar G$, and the generation-interval dispersion parameter $\kappa$ with appropriate probability distributions;
we used Gamma distributions to model values reported with confidence intervals and uniform distributions to model values reported with ranges.
For example, Study 2 estimated $\Ro = 2.92$ (95\% CI: 2.28--3.67);
we model this estimate as a Gamma distribution with a mean of 2.92 and a shape parameter of 67, which has a 95\% probability of containing a value between 2.28 and 3.67 (see Table 2 for a complete description).
We then constructed a family of parameter sets --- which include $r$, $\bar G$, and $\kappa$ --- for each study and used these in a Bayesian multilevel model to build a distribution of pooled estimates for each parameter ($\mu_r$, $\mu_G$, and $\mu_\kappa$).
The pooled estimates, which are represented as probability distributions rather than point estimates, allow us to average across different modeling approaches, while accounting for the uncertainties in the assumptions they make (see Methods).

\newcommand{\gammdist}{\mathrm{Gamma}}
\begin{table}[t]
\begin{center}
\scriptsize
\begin{tabular}{l|p{4.5cm}|p{2.5cm}|p{2.7cm}}
 & Basic reproductive\newline number \Ro & Mean generation\newline interval $\bar G$ (days) & Generation-interval\newline dispersion $\kappa$  \\
\hline
Study 1 & $\gammdist(\mathrm{mean}=2.6, \alpha=28)$ & 8.4 & 0.5 \\
\hline
Study 2 & $\gammdist(\mathrm{mean}=2.92, \alpha=67)$ & 8.4 & 0.2 \\
\hline
Study 3 & $\gammdist(\mathrm{mean}=3.8, \alpha=1400)$ & 7.6 & 0.5 \\
\hline
Study 4 & $\gammdist(\mathrm{mean}=2.2, \alpha=12)$ & $\mathrm{Uniform}(7, 14)$ & 0.5\\
\hline
Study 5 & $\gammdist(\mathrm{mean}=5.47, \alpha=54)$ & $\mathrm{Uniform}(7.6, 8.4)^\ast$ & 0.2\\
\hline
Study 6 & $\exp(r \bar G)^\dagger$ & $\mathrm{Uniform}(6, 10)$ & 0\\
\hline
\end{tabular}
\end{center}
\caption{
\textbf{Probability distributions for \Ro, $\bar G$, and $\kappa$.}
We use these probability distributions to obtain a probability distribution for the exponential growth rate $r$.
The Gamma distribution is parameterized by its mean and shape $\alpha$.
Constant values are fixed according to Table 1.
$^\ast$We do not account for this uncertainty during our recalculation of the exponential growth rate $r$ because the reported estimate of $\mathcal R_0$ and its uncertainty assumes $\bar G = 8$; the original article reports three $\mathcal R_0$ (and 95\% CIs) estimates using three different values of $\bar G$: 7.6 (MERS-like), 8 (average), and 8.4 (SARS-like).
We still account for this uncertainty in our pooled estimates ($\mu_G$).
$^\dagger$Study 6 uses the IDEA model \citep{fisman2013idea}, through which the authors effectively fit an exponential curve to the cumulative number of confirmed cases without propagating any statistical uncertainty.
Instead of modeling \Ro with a probability distribution and recalculating $r$, we use $r=0.114\,\mathrm{days}^{-1}$, which explains all uncertainty in \Ro they report, when combined with the range of $\bar G$ they consider.
}
\end{table}

\fref{assumption} compares the reported values of the exponential growth rate $r$, mean generation interval $\bar G$, and the generation-interval dispersion $\kappa$ from different studies with the pooled estimates that we calculate from our multilevel model.
We find that there is a large uncertainty associated with the underlying parameters;
many models rely on stronger assumptions that ignore these uncertainties.
Surprisingly, no studies take into account how the variation in generation intervals affects their estimates of \Ro:
all studies assumed fixed values for $\kappa$, ranging from 0 to 0.5.
Assuming fixed parameter values can lead to overly strong conclusions \citep{elderd2006uncertainty}.
It is also worth to noting that none of the six studies explicitly or implicitly assumed an exponentially distributed generation interval (i.e., $\kappa=1$), an assumption which used to be extremely common, particularly implicitly (e.g., \cite{may1987transmission, anderson1991infectious, pybus2001epidemic, ferguson2005strategies}).

@@

\fref{eff} shows how propagating uncertainty in different combinations would affect estimates and CIs for \Ro. For illustrative purposes, we use our pooled estimates, which may represent a reasonable proxy for the state of knowledge as of January 26.
Comparing the models that include only some sources of uncertainty to the ``all'' model, we see that propagating error from the growth rate (which all but one of the studies reviewed did) is absolutely crucial: the middle bar (\textbf{GI mean}), which lacks growth-rate uncertainty, is far too narrow.
Propagating error from the generation interval also has important effects.
Once these two are included, the impact of leaving out uncertainty in the dispersion is small, though noticeable, in this particular example.

@@

We also evaluate the estimates of \Ro across different studies by 
replacing their values of $r$, $\bar G$, and $\kappa$ with our pooled estimates ($\mu_r$, $\mu_G$, and $\mu_\kappa$, respectively) one at a time and recalculating the basic reproductive number \Ro (\fref{R0}).
This procedure allows us to assess the sensitivity of the estimates of \Ro across appropriate ranges of uncertainties and test how underlying assumptions drive the sensitivity.
We find that incorporating uncertainties one at a time increases the width of the confidence intervals in all but three cases.
We estimate slightly narrower confidence intervals for Study 2 and Study 6 when we use our pooled estimate of the generation-interval dispersion $\mu_\kappa$ to recalculate \Ro because they assume a narrow generation-interval distribution (compare \textbf{base} with \textbf{GI variation} in \fref{R0});
when higher values of $\kappa$ are used, their estimates of \Ro become less sensitive to the values of $r$ and $\bar G$, giving narrower confidence intervals.
We estimate narrower confidence intervals for Study 4 when we use our pooled estimate of the mean generation interval $\mu_G$ to recalculate \Ro (compare \textbf{base} with \textbf{GI mean} in \fref{R0}) because the range of uncertainty in the mean generation interval $\bar G$ they consider is much wider than the pooled range (\fref{assumption}).

Consistent with our previous observations (\fref{eff}),
we find that accounting for uncertainties in the estimate of $r$ has the largest effect on the estimates of \Ro (\fref{R0}).
For example, recalculating \Ro for Study 6 by using our pooled estimate of $r$ gives $\Ro = 3.9$ (95\% CI: 2.3--9.8), which is much wider than the uncertainty range they reported (2.0--3.1).
There are two explanations for this result.
First, even though the exponential growth rate $r$ and the mean generation interval $\bar G$ have identical effects on \Ro under the gamma approximation framework (\eref{gamma} in Methods),
$r$ has a greater overall effect on \Ro because it is associated with more uncertainty (\fref{assumption}).
Second, assuming a fixed generation interval ($\kappa=0$) makes the estimate of \Ro too sensitive to $r$ and $\bar G$ as discussed previously.

@@

Here, we focused on the estimates of \Ro that were published within a very short frame of time (January 24--26).
Although our analysis only reflects a snapshot of a fast-moving epidemic, our lessons will hold: confidence intervals must combine different sources of uncertainty. 
In fact, as epidemics progress and more data becomes available, it is likely that inferences about exponential growth rate will become more precise; thus the risk of over-confidence when uncertainty about the generation-interval distribution is neglected will become greater.

@@

We reviewed 6 modeling studies of the novel coronavirus outbreak that were published online between January 24 and January 26, 2020 (Table 1).
Five studies \citep{liuncov, majumderncov, readncov, riouncov, zhaoncov} were uploaded to pre-print servers (bioRxiv, medRxiv, and SSRN), and one report was posted on the website of Imperial College London \citep{imaincov}.
They vary widely in their statistical methods and in the amount of data used to infer \Ro.
\cite{imaincov} and \cite{riouncov} simulated branching process models and compared the predicted number of cases from their models with the estimated number of total cases by January 18.
\cite{readncov} fitted a deterministic, metapopulation Susceptible-Exposed-Infected-Recovered (SEIR) model to incidence data between January 1 and January 21 from major cities in China and other countries.
\cite{zhaoncov} and \cite{liuncov} fitted exponential growth models to incidence data up to January 22 and January 23, respectively, and inferred \Ro via the Euler-Lotka equation (\eref{euler}).
\cite{majumderncov} fitted the Incidence Decay and Exponential Adjustment (IDEA) model \citep{fisman2013idea} to incidence data up to January 26, which is equivalent to fitting an exponential growth model and assuming a fixed generation-interval distribution.

@@

\section*{Acknowledgements}

We thank Ben Bolker for providing useful comments on the manuscript.

>>>>

We evaluate six disparate models published online between January 23--26, 2020 that estimated \Ro for the 2019-nCoV outbreak \citep{imaincov, liuncov, majumderncov, readncov, riouncov, zhaoncov}. 
We use our framework to construct pooled estimates for the three key quantities: $r$, $\bar G$, and $\kappa$;
the pooled estimates reflect the uncertainties present in modeling approaches and their underlying assumptions.
We use these pooled estimates to illustrate the importance of propagating different sources of error, particularly uncertainty in both the growth rate and the generation interval. 
We also use our framework to tease apart which assumptions of these different models led to their different estimates and confidence intervals.

@@

\begin{table}[t]
\begin{center}
\scriptsize
\begin{tabular}{l|p{3.5cm}|p{2.5cm}|p{2.7cm}|l}
 & Basic reproductive\newline number \Ro & Mean generation\newline interval $\bar G$ (days) & Generation-interval\newline dispersion $\kappa$ \\
\hline
Study 1 & 1.5--3.5 & 10 & 1 & \cite{bedfordncov} \\
\hline
Study 2 & 2.5 (1.5--3.5)$^\ast$ & 8.4 & unspecified$^\dagger$ & \cite{imaincov} \\
\hline
Study 3 & 2.92 (95\% CI: 2.28--3.67) & 8.4 & 0.2 & \cite{liuncov} \\
\hline
Study 4 & 3.8 (95\% CI: 3.6--4.0) & 7.6 & 0.5 & \cite{readncov} \\
\hline
Study 5 & 2.2 (90\% CI: 1.4--3.8) & 7--14 & 0.5 & \cite{riouncov} \\
\hline
Study 6 & 5.47 (95\% CI: 4.16--7.10)$^\ddagger$ & 7.6--8.4 & 0.2 & \cite{zhaoncov} \\
\hline
Study 7 & 2.0--3.1 & 6--10 & 0 & \cite{majumderncov} \\
\hline
\end{tabular}
\end{center}
\caption{
\textbf{Reported estimates of the basic reproductive number and the assumptions about the generation-interval distributions.}
Estimates of \Ro and their assumptions about the shape of the generation interval distributions were collected from 6 studies.
$^\ast$We treat these intervals as a 95\% confidence interval in our analysis.
$^\dagger$We assume $\kappa = 0.5$ in our analysis.
$^\ddagger$The authors presented \Ro estimates under different assumptions; we use their baseline scenario in our analysis.
}
\end{table}

We gathered information on estimates of \Ro and their assumptions about the underlying generation-interval distributions from 6 articles that were published online between January 23 and January 26, 2020 (Table 1).
As most studies do not report their estimates of the exponential growth rate or the associated confidence intervals, we first recalculate the exponential growth rate that correspond their model assumptions.
We do so by modeling explicitly or implicitly reported distributions of the reproductive number \Ro, the mean generation interval $\bar G$, and the generation-interval dispersion parameter $\kappa$ with appropriate probability distributions;
we used Gamma distributions to model values reported with confidence intervals and uniform distributions to model values reported with ranges.
For example, Study 3 estimated $\Ro = 2.92$ (95\% CI: 2.28--3.67);
we model this estimate as a Gamma distribution with a mean of 2.92 and a shape parameter of 67, which has a 95\% probability of containing a value between 2.28 and 3.67 (see Table 2 for a complete description).
We then constructed a family of parameter sets --- which include $r$, $\bar G$, and $\kappa$ --- for each study and used these in a Bayesian multilevel model to build a distribution of pooled estimates for each parameter ($\mu_r$, $\mu_G$, and $\mu_\kappa$).
The pooled estimates, which are represented as probability distributions rather than point estimates, allow us to average across different modeling approaches, while accounting for the uncertainties in the assumptions they make (see Methods).

\newcommand{\gammdist}{\mathrm{Gamma}}
\begin{table}[t]
\begin{center}
\scriptsize
\begin{tabular}{l|p{4.5cm}|p{2.5cm}|p{2.7cm}}
 & Basic reproductive\newline number \Ro & Mean generation\newline interval $\bar G$ (days) & Generation-interval\newline dispersion $\kappa$ \\
\hline
Study 1 & $\mathrm{Uniform}(1.5, 3.5)$ & 10 & 1 \\
\hline
Study 2 & $\gammdist(\mathrm{mean}=2.6, \alpha=28)$ & 8.4 & 0.5 \\
\hline
Study 3 & $\gammdist(\mathrm{mean}=2.92, \alpha=67)$ & 8.4 & 0.2 \\
\hline
Study 4 & $\gammdist(\mathrm{mean}=3.8, \alpha=1400)$ & 7.6 & 0.5 \\
\hline
Study 5 & $\gammdist(\mathrm{mean}=2.2, \alpha=12)$ & $\mathrm{Uniform}(7, 14)$ & 0.5\\
\hline
Study 6 & $\gammdist(\mathrm{mean}=5.47, \alpha=54)$ & $\mathrm{Uniform}(7.6, 8.4)^\ast$ & 0.2\\
\hline
Study 7 & $\exp(r \bar G)^\dagger$ & $\mathrm{Uniform}(6, 10)$ & 0\\
\hline
\end{tabular}
\end{center}
\caption{
\textbf{Probability distributions for \Ro, $\bar G$, and $\kappa$.}
We use these probability distributions to obtain a probability distribution for the exponential growth rate $r$.
The Gamma distribution is parameterized by its mean and shape $\alpha$.
Constant values are fixed according to Table 1.
$^\ast$We do not account for this uncertainty during our recalculation of the exponential growth rate $r$ because the reported estimate of $\mathcal R_0$ and its uncertainty assumes $\bar G = 8$; the original article reports three $\mathcal R_0$ (and 95\% CIs) estimates using three different values of $\bar G$: 7.6 (MERS-like), 8 (average), and 8.4 (SARS-like).
We still account for this uncertainty in our pooled estimates ($\mu_G$).
$^\dagger$Study 6 uses the IDEA model \citep{fisman2013idea}, through which the authors effectively fit an exponential curve to the cumulative number of confirmed cases without propagating any statistical uncertainty.
Instead of modeling \Ro with a probability distribution and recalculating $r$, we use $r=0.114\,\mathrm{days}^{-1}$, which explains all uncertainty in \Ro they report, when combined with the range of $\bar G$ they consider.
}
\end{table}

\fref{assumption} compares the reported values of the exponential growth rate $r$, mean generation interval $\bar G$, and the generation-interval dispersion $\kappa$ from different studies with the pooled estimates that we calculate from our multilevel model.
We find that there is a large uncertainty associated with the underlying parameters;
many models rely on stronger assumptions that ignore these uncertainties.
Surprisingly, no studies take into account how the variation in generation intervals affects their estimates of \Ro:
all studies assumed fixed values for $\kappa$, ranging from 0 to 0.5.
Assuming fixed parameter values can lead to overly strong conclusions \citep{elderd2006uncertainty}.
It is also worth noting that none but one studies explicitly or implicitly assumed an exponentially distributed generation interval (i.e., $\kappa=1$), an assumption which used to be extremely common, particularly implicitly (e.g., \cite{may1987transmission, anderson1991infectious, pybus2001epidemic, ferguson2005strategies}).

@@

\fref{eff} shows how propagating uncertainty in different combinations would affect estimates and CIs for \Ro. For illustrative purposes, we use our pooled estimates, which may represent a reasonable proxy for the state of knowledge as of January 23--26.
Comparing the models that include only some sources of uncertainty to the ``all'' model, we see that propagating error from the growth rate (which all but one of the studies reviewed did) is absolutely crucial: the middle bar (\textbf{GI mean}), which lacks growth-rate uncertainty, is far too narrow.
Propagating error from the generation interval also has important effects.
Once these two are included, the impact of leaving out uncertainty in the dispersion is small, though noticeable, in this particular example.

@@

We also evaluate the estimates of \Ro across different studies by 
replacing their values of $r$, $\bar G$, and $\kappa$ with our pooled estimates ($\mu_r$, $\mu_G$, and $\mu_\kappa$, respectively) one at a time and recalculating the basic reproductive number \Ro (\fref{R0}).
This procedure allows us to assess the sensitivity of the estimates of \Ro across appropriate ranges of uncertainties and test how underlying assumptions drive the sensitivity.
We find that incorporating uncertainties one at a time increases the width of the confidence intervals in all but three cases.
We estimate slightly narrower confidence intervals for Study 3 and Study 7 when we use our pooled estimate of the generation-interval dispersion $\mu_\kappa$ to recalculate \Ro because they assume a narrow generation-interval distribution (compare \textbf{base} with \textbf{GI variation} in \fref{R0});
when higher values of $\kappa$ are used, their estimates of \Ro become less sensitive to the values of $r$ and $\bar G$, giving narrower confidence intervals.
We estimate narrower confidence intervals for Study 4 when we use our pooled estimate of the mean generation interval $\mu_G$ to recalculate \Ro (compare \textbf{base} with \textbf{GI mean} in \fref{R0}) because the range of uncertainty in the mean generation interval $\bar G$ they consider is much wider than the pooled range (\fref{assumption}).

Consistent with our previous observations (\fref{eff}),
we find that accounting for uncertainties in the estimate of $r$ has the largest effect on the estimates of \Ro (\fref{R0}).
For example, recalculating \Ro for Study 7 by using our pooled estimate of $r$ gives $\Ro = 3.9$ (95\% CI: 2.3--9.8), which is much wider than the uncertainty range they reported (2.0--3.1).
There are two explanations for this result.
First, even though the exponential growth rate $r$ and the mean generation interval $\bar G$ have identical effects on \Ro under the gamma approximation framework (\eref{gamma} in Methods),
$r$ has a greater overall effect on \Ro because it is associated with more uncertainty (\fref{assumption}).
Second, assuming a fixed generation interval ($\kappa=0$) makes the estimate of \Ro too sensitive to $r$ and $\bar G$ as discussed previously.

@@

Here, we focused on the estimates of \Ro that were published within a very short frame of time (January 23--26).
Although our analysis only reflects a snapshot of a fast-moving epidemic, our lessons will hold: confidence intervals must combine different sources of uncertainty. 
In fact, as epidemics progress and more data becomes available, it is likely that inferences about exponential growth rate will become more precise; thus the risk of over-confidence when uncertainty about the generation-interval distribution is neglected will become greater.

@@

We reviewed 6 modeling studies of the novel coronavirus outbreak that were published online between January 23 and January 26, 2020 (Table 1).
Five studies \citep{liuncov, majumderncov, readncov, riouncov, zhaoncov} were uploaded to pre-print servers (bioRxiv, medRxiv, and SSRN), and one report was posted on the website of Imperial College London \citep{imaincov}.
They vary widely in their statistical methods and in the amount of data used to infer \Ro.
\cite{imaincov} and \cite{riouncov} simulated branching process models and compared the predicted number of cases from their models with the estimated number of total cases by January 18.
\cite{readncov} fitted a deterministic, metapopulation Susceptible-Exposed-Infected-Recovered (SEIR) model to incidence data between January 1 and January 21 from major cities in China and other countries.
\cite{zhaoncov} and \cite{liuncov} fitted exponential growth models to incidence data up to January 22 and January 23, respectively, and inferred \Ro via the Euler-Lotka equation (\eref{euler}).
\cite{majumderncov} fitted the Incidence Decay and Exponential Adjustment (IDEA) model \citep{fisman2013idea} to incidence data up to January 26, which is equivalent to fitting an exponential growth model and assuming a fixed generation-interval distribution.

@@
